---
title: "Treatment of colonic and rectal adenomas with sulindac in familial adenomatous polyposis"
author: "Daniel Herrera"
format: html
---

## Data source

This data set is from a study published in 1993 in the New England Journal of Medicine: F. M. Giardiello, S. R. Hamilton, A. J. Krush, S. Piantadosi, L. M. Hylind, P. Celano, S. V. Booker, C. R. Robinson and G. J. A. Offerhaus (1993), Treatment of colonic and rectal adenomas with sulindac in familial adenomatous polyposis. New England Journal of Medicine, 328(18), 1313–1316.

It is part of the medicaldata package <https://github.com/higgi13425/medicaldata/tree/master/data>, "polyps" data set.

## **Background**

Familial adenomatous polyposis (FAP) is an inherited condition caused by mutations in the APC (Adenomatous Polyposis Coli) gene that leads to early and frequent formation of precancerous polyps of the colon at a young age, and invariably leads to the development of colon cancer at a young age. Early, frequent surveillance colonoscopy and polyp removal is helpful, but this study examined whether there is a beneficial effect of preventive medical therapy with the nonsteroidal pain reliever, sulindac, versus placebo in a RCT vs placebo in 22 participants, with polyp number measured (via colonoscopy) at baseline, 3 months, and 12 months after starting the study drug.

## **Study Design**

Prospective randomized controlled trial of sulindac vs. placebo in the prevention of colonic polyps in FAP.

### **Subjects & Variables**

22 participants were assigned to 1 of 2 treatments, sulindac or placebo. The initial sample consisted of 22 people with known FAP. n = 22 subjects (2 dropped out between month 3 and 12)

7 variables: participant_id, sex, age, baseline (number of polyps at baseline visit), treatment (sulindac or placebo), number3m (number of polyps at 3 months), and number6m (number of polyps at 6 months).

## Results

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
